---
title: "Dental Age Continuity"
format: html
vignette: >
  %\VignetteIndexEntry{Dental Age Continuity}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| message: false
library(AgeFromDentition)
library(dplyr)
library(tibble)
library(ggplot2)
library(gt)
library(tidyr)
```

Dental age estimates generally increase as tooth stages advance. This
vignette constructs fine-scale example progressions for females and
males, starting with only the canine scored and gradually adding teeth,
ending with all teeth at `A.c` except M3 at `A.5`.

## Female progression

```{r}
#| label: female-data
female <- tribble(
    ~Sex , ~Canine , ~P3     , ~P4     , ~M1    , ~M2     , ~M3     ,
    "F"  , "C.oc"  , NA      , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.5"  , NA      , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.5"  , "C.i"   , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.5"  , "C.co"  , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.75" , "C.oc"  , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.75" , "C.oc"  , NA      , "Cr.c" , NA      , NA      ,
    "F"  , "Cr.75" , "Cr.5"  , NA      , "R.i"  , NA      , NA      ,
    "F"  , "Cr.75" , "Cr.5"  , NA      , "R.i"  , "C.i"   , NA      ,
    "F"  , "Cr.75" , "Cr.5"  , "C.i"   , "Cl.i" , "C.i"   , NA      ,
    "F"  , "Cr.c"  , "Cr.75" , "C.co"  , "R.25" , "C.co"  , NA      ,
    "F"  , "Cr.c"  , "Cr.c"  , "C.oc"  , "R.25" , "C.oc"  , NA      ,
    "F"  , "Cr.c"  , "Cr.c"  , "Cr.5"  , "R.5"  , "Cr.5"  , NA      ,
    "F"  , "R.i"   , "Cr.c"  , "Cr.75" , "R.75" , "Cr.75" , NA      ,
    "F"  , "R.i"   , "R.i"   , "Cr.c"  , "R.75" , "Cr.c"  , NA      ,
    "F"  , "R.25"  , "R.i"   , "R.i"   , "R.75" , "R.i"   , NA      ,
    "F"  , "R.25"  , "R.25"  , "R.i"   , "R.c"  , "Cl.i"  , NA      ,
    "F"  , "R.5"   , "R.25"  , "R.25"  , "R.c"  , "R.25"  , NA      ,
    "F"  , "R.5"   , "R.5"   , "R.25"  , "A.5"  , "R.25"  , NA      ,
    "F"  , "R.75"  , "R.5"   , "R.5"   , "A.c"  , "R.5"   , NA      ,
    "F"  , "R.75"  , "R.5"   , "R.5"   , "A.c"  , "R.5"   , "C.i"   ,
    "F"  , "R.c"   , "R.75"  , "R.75"  , "A.c"  , "R.75"  , "C.co"  ,
    "F"  , "A.5"   , "R.c"   , "R.c"   , "A.c"  , "R.c"   , "C.oc"  ,
    "F"  , "A.c"   , "A.5"   , "A.5"   , "A.c"  , "A.5"   , "Cr.5"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cr.75" ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cr.c"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.i"   ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cl.i"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.25"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.5"   ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.75"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.c"   ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "A.5"   ,
)
```

```{r}
#| label: female-estimates
female_results <- purrr::map_dfr(seq_len(nrow(female)), \(i) {
    est <- estimate_dental_age(female[i, ], verbose = FALSE)
    info <- ac_info(est)
    tibble(
        row = i,
        dental_age = est[["dental_age"]],
        ci_lower = est[["ci_lower"]],
        ci_upper = est[["ci_upper"]],
        n_estimable = info$n_estimable,
        compatibility = info$compatibility
    )
})

Females <- female |>
    mutate(row = row_number()) |>
    left_join(female_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(c(dental_age, ci_lower, ci_upper), \(x) round(x, 2))) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Females |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(label = "95% CI", columns = c(ci_lower, ci_upper))
```

```{r}
Females |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_classic()
```

## Female progression without M3

The same progression, but M3 is never scored. Without M3, the late
rows where all other teeth have reached `A.c` produce no point
estimate — only a completion threshold.

```{r}
#| label: female-no-m3-data
female_no_m3 <- female |>
    mutate(M3 = NA)
```

```{r}
#| label: female-no-m3-estimates
female_no_m3_results <- purrr::map_dfr(
    seq_len(nrow(female_no_m3)),
    \(i) {
        est <- estimate_dental_age(
            female_no_m3[i, ],
            verbose = FALSE
        )
        info <- ac_info(est)
        tibble(
            row = i,
            dental_age = est[["dental_age"]],
            ci_lower = est[["ci_lower"]],
            ci_upper = est[["ci_upper"]],
            n_estimable = info$n_estimable,
            compatibility = info$compatibility
        )
    }
)

Females_no_M3 <- female_no_m3 |>
    mutate(row = row_number()) |>
    left_join(female_no_m3_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(
        c(dental_age, ci_lower, ci_upper),
        \(x) round(x, 2)
    )) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Females_no_M3 |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(
        label = "95% CI",
        columns = c(ci_lower, ci_upper)
    )
```

```{r}
Females_no_M3 |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_classic()
```

### Comparison: with and without M3

```{r}
#| label: female-comparison
#| echo: false
bind_rows(
    Females |>
        drop_na(dental_age) |>
        mutate(M3_status = "With M3"),
    Females_no_M3 |>
        drop_na(dental_age) |>
        mutate(M3_status = "Without M3")
) |>
    ggplot(aes(x = row, y = dental_age, color = M3_status)) +
    geom_line(linewidth = 1.5) +
    labs(
        x = "Row",
        y = "Dental Age",
        title = "Female",
        color = NULL
    ) +
    theme_classic()
```


## Male progression

```{r}
#| label: male-data
male <- tribble(
    ~Sex , ~Canine , ~P3     , ~P4     , ~M1    , ~M2     , ~M3     ,
    "M"  , "C.oc"  , NA      , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.5"  , NA      , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.5"  , "C.i"   , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.5"  , "C.co"  , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.75" , "C.oc"  , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.75" , "C.oc"  , NA      , "Cr.c" , NA      , NA      ,
    "M"  , "Cr.75" , "Cr.5"  , NA      , "R.i"  , NA      , NA      ,
    "M"  , "Cr.75" , "Cr.5"  , NA      , "R.i"  , "C.i"   , NA      ,
    "M"  , "Cr.75" , "Cr.5"  , "C.i"   , "Cl.i" , "C.i"   , NA      ,
    "M"  , "Cr.c"  , "Cr.75" , "C.co"  , "R.25" , "C.co"  , NA      ,
    "M"  , "Cr.c"  , "Cr.c"  , "C.oc"  , "R.25" , "C.oc"  , NA      ,
    "M"  , "Cr.c"  , "Cr.c"  , "Cr.5"  , "R.5"  , "Cr.5"  , NA      ,
    "M"  , "R.i"   , "Cr.c"  , "Cr.75" , "R.75" , "Cr.75" , NA      ,
    "M"  , "R.i"   , "R.i"   , "Cr.c"  , "R.75" , "Cr.c"  , NA      ,
    "M"  , "R.25"  , "R.i"   , "R.i"   , "R.75" , "R.i"   , NA      ,
    "M"  , "R.25"  , "R.25"  , "R.i"   , "R.c"  , "Cl.i"  , NA      ,
    "M"  , "R.5"   , "R.25"  , "R.25"  , "R.c"  , "R.25"  , NA      ,
    "M"  , "R.5"   , "R.5"   , "R.25"  , "A.5"  , "R.25"  , NA      ,
    "M"  , "R.75"  , "R.5"   , "R.5"   , "A.c"  , "R.5"   , NA      ,
    "M"  , "R.75"  , "R.5"   , "R.5"   , "A.c"  , "R.5"   , "C.i"   ,
    "M"  , "R.c"   , "R.75"  , "R.75"  , "A.c"  , "R.75"  , "C.co"  ,
    "M"  , "A.5"   , "R.c"   , "R.c"   , "A.c"  , "R.c"   , "C.oc"  ,
    "M"  , "A.c"   , "A.5"   , "A.5"   , "A.c"  , "A.5"   , "Cr.5"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cr.75" ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cr.c"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.i"   ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cl.i"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.25"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.5"   ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.75"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.c"   ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "A.5"   ,
)
```

```{r}
#| label: male-estimates
male_results <- purrr::map_dfr(seq_len(nrow(male)), \(i) {
    est <- estimate_dental_age(male[i, ], verbose = FALSE)
    info <- ac_info(est)
    tibble(
        row = i,
        dental_age = est[["dental_age"]],
        ci_lower = est[["ci_lower"]],
        ci_upper = est[["ci_upper"]],
        n_estimable = info$n_estimable,
        compatibility = info$compatibility
    )
})

Males <- male |>
    mutate(row = row_number()) |>
    left_join(male_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(c(dental_age, ci_lower, ci_upper), \(x) round(x, 2))) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Males |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(label = "95% CI", columns = c(ci_lower, ci_upper))
```

```{r}
Males |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_classic()
```

## Male progression without M3

```{r}
#| label: male-no-m3-data
male_no_m3 <- male |>
    mutate(M3 = NA)
```

```{r}
#| label: male-no-m3-estimates
male_no_m3_results <- purrr::map_dfr(
    seq_len(nrow(male_no_m3)),
    \(i) {
        est <- estimate_dental_age(
            male_no_m3[i, ],
            verbose = FALSE
        )
        info <- ac_info(est)
        tibble(
            row = i,
            dental_age = est[["dental_age"]],
            ci_lower = est[["ci_lower"]],
            ci_upper = est[["ci_upper"]],
            n_estimable = info$n_estimable,
            compatibility = info$compatibility
        )
    }
)

Males_no_M3 <- male_no_m3 |>
    mutate(row = row_number()) |>
    left_join(male_no_m3_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(
        c(dental_age, ci_lower, ci_upper),
        \(x) round(x, 2)
    )) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Males_no_M3 |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(
        label = "95% CI",
        columns = c(ci_lower, ci_upper)
    )
```

```{r}
Males_no_M3 |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_classic()
```

### Comparison: with and without M3

```{r}
#| label: male-comparison
#| echo: false
bind_rows(
    Males |>
        drop_na(dental_age) |>
        mutate(M3_status = "With M3"),
    Males_no_M3 |>
        drop_na(dental_age) |>
        mutate(M3_status = "Without M3")
) |>
    ggplot(aes(x = row, y = dental_age, color = M3_status)) +
    geom_line(linewidth = 1.5) +
    labs(
        x = "Row",
        y = "Dental Age",
        title = "Male",
        color = NULL
    ) +
    theme_classic()
```

## Notes

- Rows 1--2 use only the canine. With a single tooth, the estimate
  tracks that tooth's reference age directly.
- P3 first appears at row 3, M1 at row 6 (already at `Cr.c` because
  it develops earliest), M2 at row 8, P4 at row 9.
- M3 first appears at row 20. For females, late M3 stages (`R.5`
  through `A.5`) have no published log-normal parameters and are
  excluded from the age estimate.
- From row 19 on, M1 reaches `A.c` and shifts from the age
  estimate to the completion threshold. The `compatibility` column
  shows whether the two agree.
- Rows 24--32 have all teeth at `A.c` except M3; only M3 contributes
  to the age estimate. For females, M3 has no parameters for `R.5`
  through `A.5`, so the final rows produce no point estimate --- only
  the completion threshold. For males, all M3 stages have parameters.
- Row 32 has all teeth `A.c` except M3 at `A.5`.
- Small dips in dental age occur when a new tooth is added at an
  early stage (e.g., M3 at `C.i` in row 20) or when several teeth
  transition to `Ac` at once (row 23 to 24).
- In the M3-absent sections, rows 1--23 are identical to the full
  progressions (M3 was `NA` in those rows anyway). From row 24 on,
  all scored teeth are at `A.c` and there is no point estimate —
  only the completion threshold.
