---
title: "Dental Age Continuity"
format: html
vignette: >
  %\VignetteIndexEntry{Dental Age Continuity}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| message: false
library(AgeFromDentition)
library(dplyr)
library(tibble)
library(ggplot2)
library(gt)
library(tidyr)
```

Dental age estimates generally increase as tooth stages advance. This
vignette constructs fine-scale example progressions for females and
males, starting with only the canine scored and gradually adding teeth,
ending with all teeth at `A.c` except M3 at `A.5`.

## Female progression

```{r}
#| label: female-data
female <- tribble(
    ~Sex , ~Canine , ~P3     , ~P4     , ~M1    , ~M2     , ~M3     ,
    "F"  , "C.oc"  , NA      , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.5"  , NA      , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.5"  , "C.i"   , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.5"  , "C.co"  , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.75" , "C.oc"  , NA      , NA     , NA      , NA      ,
    "F"  , "Cr.75" , "C.oc"  , NA      , "R.i"  , NA      , NA      ,
    "F"  , "Cr.75" , "Cr.5"  , NA      , "Cl.i" , NA      , NA      ,
    "F"  , "Cr.75" , "Cr.5"  , NA      , "Cl.i" , "C.i"   , NA      ,
    "F"  , "Cr.75" , "Cr.5"  , "C.i"   , "R.25" , "C.i"   , NA      ,
    "F"  , "Cr.c"  , "Cr.75" , "C.co"  , "R.25" , "C.co"  , NA      ,
    "F"  , "Cr.c"  , "Cr.c"  , "C.oc"  , "R.5"  , "C.oc"  , NA      ,
    "F"  , "Cr.c"  , "Cr.c"  , "Cr.5"  , "R.5"  , "Cr.5"  , NA      ,
    "F"  , "R.i"   , "Cr.c"  , "Cr.75" , "R.75" , "Cr.75" , NA      ,
    "F"  , "R.i"   , "R.i"   , "Cr.c"  , "R.75" , "Cr.c"  , NA      ,
    "F"  , "R.25"  , "R.i"   , "R.i"   , "R.75" , "R.i"   , NA      ,
    "F"  , "R.25"  , "R.25"  , "R.i"   , "R.c"  , "Cl.i"  , NA      ,
    "F"  , "R.5"   , "R.25"  , "R.25"  , "R.c"  , "R.25"  , NA      ,
    "F"  , "R.5"   , "R.5"   , "R.25"  , "A.5"  , "R.25"  , NA      ,
    "F"  , "R.75"  , "R.5"   , "R.5"   , "A.c"  , "R.5"   , NA      ,
    "F"  , "R.75"  , "R.5"   , "R.5"   , "A.c"  , "R.5"   , "C.oc"  ,
    "F"  , "R.c"   , "R.75"  , "R.75"  , "A.c"  , "R.75"  , "Cr.5"  ,
    "F"  , "A.5"   , "R.c"   , "R.c"   , "A.c"  , "R.c"   , "Cr.75" ,
    "F"  , "A.c"   , "A.5"   , "A.5"   , "A.c"  , "A.5"   , "Cr.c"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cr.c"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.i"   ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.i"   ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cl.i"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.25"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.5"   ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.75"  ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.c"   ,
    "F"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "A.5"   ,
)
```

```{r}
#| label: female-estimates
female_results <- purrr::map_dfr(seq_len(nrow(female)), \(i) {
    est <- estimate_dental_age(female[i, ], verbose = FALSE)
    info <- ac_info(est)
    tibble(
        row = i,
        dental_age = est[["dental_age"]],
        ci_lower = est[["ci_lower"]],
        ci_upper = est[["ci_upper"]],
        n_estimable = info$n_estimable,
        compatibility = info$compatibility
    )
})

Females <- female |>
    mutate(row = row_number()) |>
    left_join(female_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(c(dental_age, ci_lower, ci_upper), \(x) round(x, 2))) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Females |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(label = "95% CI", columns = c(ci_lower, ci_upper))
```

```{r}
Females |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_minimal()
```

## Female progression without M3

The same progression, but M3 is never scored. Without M3, the late
rows where all other teeth have reached `A.c` produce no point
estimate — only a completion threshold.

```{r}
#| label: female-no-m3-data
female_no_m3 <- female |>
    mutate(M3 = NA)
```

```{r}
#| label: female-no-m3-estimates
female_no_m3_results <- purrr::map_dfr(
    seq_len(nrow(female_no_m3)),
    \(i) {
        est <- estimate_dental_age(
            female_no_m3[i, ],
            verbose = FALSE
        )
        info <- ac_info(est)
        tibble(
            row = i,
            dental_age = est[["dental_age"]],
            ci_lower = est[["ci_lower"]],
            ci_upper = est[["ci_upper"]],
            n_estimable = info$n_estimable,
            compatibility = info$compatibility
        )
    }
)

Females_no_M3 <- female_no_m3 |>
    mutate(row = row_number()) |>
    left_join(female_no_m3_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(
        c(dental_age, ci_lower, ci_upper),
        \(x) round(x, 2)
    )) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Females_no_M3 |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(
        label = "95% CI",
        columns = c(ci_lower, ci_upper)
    )
```

```{r}
Females_no_M3 |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_minimal()
```

### Comparison: with and without M3

```{r}
#| label: female-comparison
#| echo: false
bind_rows(
    Females |>
        drop_na(dental_age) |>
        mutate(M3_status = "With M3"),
    Females_no_M3 |>
        drop_na(dental_age) |>
        mutate(M3_status = "Without M3")
) |>
    ggplot(aes(x = row, y = dental_age, color = M3_status)) +
    geom_line(linewidth = 1.5) +
    scale_color_manual(values = c("navy", "firebrick")) +
    labs(
        x = "Row",
        y = "Dental Age",
        title = "Female",
        color = NULL
    ) +
    theme_minimal()
```


## Male progression

```{r}
#| label: male-data
male <- tribble(
    ~Sex , ~Canine , ~P3     , ~P4     , ~M1    , ~M2     , ~M3     ,
    "M"  , "C.oc"  , NA      , NA      , NA     , NA      , NA      ,
    "M"  , "C.oc"  , NA      , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.5"  , "C.i"   , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.5"  , "C.i"   , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.5"  , "C.co"  , NA      , NA     , NA      , NA      ,
    "M"  , "Cr.5"  , "C.oc"  , NA      , "Cr.c" , NA      , NA      ,
    "M"  , "Cr.75" , "C.oc"  , NA      , "R.i"  , NA      , NA      ,
    "M"  , "Cr.75" , "Cr.5"  , NA      , "R.i"  , "C.i"   , NA      ,
    "M"  , "Cr.75" , "Cr.5"  , "C.i"   , "Cl.i" , "C.i"   , NA      ,
    "M"  , "Cr.75" , "Cr.5"  , "C.co"  , "R.25" , "C.co"  , NA      ,
    "M"  , "Cr.75" , "Cr.75" , "C.oc"  , "R.25" , "C.oc"  , NA      ,
    "M"  , "Cr.c"  , "Cr.75" , "Cr.5"  , "R.5"  , "Cr.5"  , NA      ,
    "M"  , "Cr.c"  , "Cr.c"  , "Cr.5"  , "R.75" , "Cr.5"  , NA      ,
    "M"  , "Cr.c"  , "Cr.c"  , "Cr.75" , "R.75" , "Cr.75" , NA      ,
    "M"  , "R.i"   , "Cr.c"  , "Cr.c"  , "R.75" , "Cr.c"  , NA      ,
    "M"  , "R.i"   , "R.i"   , "R.i"   , "R.c"  , "R.i"   , NA      ,
    "M"  , "R.25"  , "R.i"   , "R.i"   , "R.c"  , "Cl.i"  , NA      ,
    "M"  , "R.25"  , "R.25"  , "R.25"  , "R.c"  , "R.25"  , NA      ,
    "M"  , "R.5"   , "R.25"  , "R.25"  , "A.5"  , "R.25"  , "C.i"   ,
    "M"  , "R.5"   , "R.5"   , "R.5"   , "A.c"  , "R.5"   , "C.co"  ,
    "M"  , "R.75"  , "R.5"   , "R.5"   , "A.c"  , "R.5"   , "C.oc"  ,
    "M"  , "R.75"  , "R.75"  , "R.75"  , "A.c"  , "R.75"  , "Cr.5"  ,
    "M"  , "R.c"   , "R.75"  , "R.75"  , "A.c"  , "R.75"  , "Cr.75" ,
    "M"  , "A.5"   , "R.c"   , "R.c"   , "A.c"  , "R.c"   , "Cr.c"  ,
    "M"  , "A.c"   , "A.5"   , "A.5"   , "A.c"  , "A.5"   , "R.i"   ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "Cl.i"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.25"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.5"   ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.75"  ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "R.c"   ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "A.5"   ,
    "M"  , "A.c"   , "A.c"   , "A.c"   , "A.c"  , "A.c"   , "A.5"   ,
)
```

```{r}
#| label: male-estimates
male_results <- purrr::map_dfr(seq_len(nrow(male)), \(i) {
    est <- estimate_dental_age(male[i, ], verbose = FALSE)
    info <- ac_info(est)
    tibble(
        row = i,
        dental_age = est[["dental_age"]],
        ci_lower = est[["ci_lower"]],
        ci_upper = est[["ci_upper"]],
        n_estimable = info$n_estimable,
        compatibility = info$compatibility
    )
})

Males <- male |>
    mutate(row = row_number()) |>
    left_join(male_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(c(dental_age, ci_lower, ci_upper), \(x) round(x, 2))) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Males |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(label = "95% CI", columns = c(ci_lower, ci_upper))
```

```{r}
Males |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_minimal()
```

## Male progression without M3

```{r}
#| label: male-no-m3-data
male_no_m3 <- male |>
    mutate(M3 = NA)
```

```{r}
#| label: male-no-m3-estimates
male_no_m3_results <- purrr::map_dfr(
    seq_len(nrow(male_no_m3)),
    \(i) {
        est <- estimate_dental_age(
            male_no_m3[i, ],
            verbose = FALSE
        )
        info <- ac_info(est)
        tibble(
            row = i,
            dental_age = est[["dental_age"]],
            ci_lower = est[["ci_lower"]],
            ci_upper = est[["ci_upper"]],
            n_estimable = info$n_estimable,
            compatibility = info$compatibility
        )
    }
)

Males_no_M3 <- male_no_m3 |>
    mutate(row = row_number()) |>
    left_join(male_no_m3_results, by = "row") |>
    select(-Sex, -row) |>
    mutate(across(
        c(dental_age, ci_lower, ci_upper),
        \(x) round(x, 2)
    )) |>
    mutate(row = row_number()) |>
    relocate(row)
```

```{r}
Males_no_M3 |>
    gt() |>
    sub_missing() |>
    cols_label(
        row = "Row",
        dental_age = "Age",
        ci_lower = "Lower",
        ci_upper = "Upper",
        n_estimable = "n",
        compatibility = "Compat."
    ) |>
    tab_spanner(label = "Scores", columns = Canine:M3) |>
    tab_spanner(
        label = "95% CI",
        columns = c(ci_lower, ci_upper)
    )
```

```{r}
Males_no_M3 |>
    drop_na(dental_age) |>
    ggplot(aes(x = row, y = dental_age)) +
    geom_line(linewidth = 1.5) +
    labs(x = "Row", y = "Dental Age") +
    theme_minimal()
```

### Comparison: with and without M3

```{r}
#| label: male-comparison
#| echo: false
bind_rows(
    Males |>
        drop_na(dental_age) |>
        mutate(M3_status = "With M3"),
    Males_no_M3 |>
        drop_na(dental_age) |>
        mutate(M3_status = "Without M3")
) |>
    ggplot(aes(x = row, y = dental_age, color = M3_status)) +
    geom_line(linewidth = 1.5) +
    scale_color_manual(values = c("navy", "firebrick")) +
    labs(
        x = "Row",
        y = "Dental Age",
        title = "Male",
        color = NULL
    ) +
    theme_minimal()
```

## Sex comparison

```{r}
#| label: sex-comparison
#| echo: false
#| fig-width: 10
#| fig-height: 5
bind_rows(
    Females |>
        drop_na(dental_age) |>
        mutate(Sex = "Female", M3_status = "With M3"),
    Females_no_M3 |>
        drop_na(dental_age) |>
        mutate(Sex = "Female", M3_status = "Without M3"),
    Males |>
        drop_na(dental_age) |>
        mutate(Sex = "Male", M3_status = "With M3"),
    Males_no_M3 |>
        drop_na(dental_age) |>
        mutate(Sex = "Male", M3_status = "Without M3")
) |>
    ggplot(aes(x = row, y = dental_age, color = Sex)) +
    geom_line(linewidth = 1.5) +
    facet_wrap(~M3_status) +
    scale_color_manual(values = c("navy", "firebrick")) +
    labs(
        x = "Row",
        y = "Dental Age",
        color = NULL
    ) +
    theme_minimal()
```


## Notes

- Rows 1--2 use only the canine. With a single tooth, the estimate
  tracks that tooth's reference age directly.
- Both sexes introduce teeth in the same order: P3 at row 3, M1
  at row 6 (already at `Cr.c` because it develops earliest), M2
  at row 8, P4 at row 9.
- The male progression uses stages that lag behind the female for
  the canine (~1 stage) and P3 (~1 stage), reflecting that males
  develop these teeth later. M1 and P4 are minimally offset. M3
  is introduced one row earlier for males (row 19 vs 20),
  reflecting slightly earlier M3 development in males.
- For females, M3 first appears at row 20; for males, at row 19.
  For females, late M3 stages (`R.5` through `A.5`) have no
  published log-normal parameters and are excluded from the age
  estimate. For males, all M3 stages have parameters.
- M1 reaches `A.c` at row 19 for females and row 20 for males,
  shifting from the age estimate to the completion threshold. The
  `compatibility` column shows whether the two agree.
- Small dips in dental age occur when a new tooth is added at an
  early stage (e.g., M3 at `C.i`) or when several teeth transition
  to `A.c` at once.
