---
title: "Dental Age Continuity"
format: html
vignette: >
  %\VignetteIndexEntry{Dental Age Continuity}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| message: false
library(AgeFromDentition)
library(dplyr)
library(tibble)
```

Dental age estimates should increase monotonically as tooth stages
advance. This vignette constructs example progressions for females and
males, starting with only the canine scored and gradually adding teeth
at developmentally appropriate stages, ending with all teeth at `Ac`
except M3 at `A.5`.

## Female progression

```{r}
#| label: female-data
female <- tribble(
  ~Sex, ~Canine, ~P3,    ~P4,    ~M1,    ~M2,    ~M3,
  "F",  "C.oc",  NA,     NA,     NA,     NA,     NA,
  "F",  "Cr.75", NA,     NA,     NA,     NA,     NA,
  "F",  "Cr.c",  "C.co", NA,     NA,     NA,     NA,
  "F",  "R.i",   "Cr.5", NA,     "Cr.c", NA,     NA,
  "F",  "R.i",   "Cr.5", "C.co", "Cr.c", NA,     NA,
  "F",  "R.25",  "Cr.c", "Cr.5", "R.i",  "C.co", NA,
  "F",  "R.5",   "R.i",  "Cr.c", "Cl.i", "Cr.5", NA,
  "F",  "R.75",  "R.25", "R.i",  "R.5",  "Cr.c", NA,
  "F",  "R.c",   "R.5",  "R.25", "R.c",  "R.25", "C.i",
  "F",  "A.5",   "R.75", "R.5",  "A.5",  "R.5",  "C.co",
  "F",  "Ac",    "R.c",  "R.75", "Ac",   "R.75", "Cr.5",
  "F",  "Ac",    "A.5",  "R.c",  "Ac",   "R.c",  "Cr.c",
  "F",  "Ac",    "Ac",   "A.5",  "Ac",   "A.5",  "R.i",
  "F",  "Ac",    "Ac",   "Ac",   "Ac",   "Ac",   "R.25",
  "F",  "Ac",    "Ac",   "Ac",   "Ac",   "Ac",   "A.5",
)
```

```{r}
#| label: female-estimates
female_results <- purrr::map_dfr(seq_len(nrow(female)), \(i) {
  est <- estimate_dental_age(female[i, ], verbose = FALSE)
  info <- ac_info(est)
  tibble(
    row = i,
    dental_age = est[["dental_age"]],
    ci_lower = est[["ci_lower"]],
    ci_upper = est[["ci_upper"]],
    n_estimable = info$n_estimable,
    compatibility = info$compatibility
  )
})

female |>
  mutate(row = row_number()) |>
  left_join(female_results, by = "row") |>
  select(-Sex, -row) |>
  mutate(across(c(dental_age, ci_lower, ci_upper), \(x) round(x, 2)))
```

## Male progression

```{r}
#| label: male-data
male <- tribble(
  ~Sex, ~Canine, ~P3,    ~P4,    ~M1,    ~M2,    ~M3,
  "M",  "C.oc",  NA,     NA,     NA,     NA,     NA,
  "M",  "Cr.75", NA,     NA,     NA,     NA,     NA,
  "M",  "Cr.c",  "C.co", NA,     NA,     NA,     NA,
  "M",  "R.i",   "Cr.5", NA,     "Cr.c", NA,     NA,
  "M",  "R.i",   "Cr.5", "C.co", "Cr.c", NA,     NA,
  "M",  "R.25",  "Cr.c", "Cr.5", "R.i",  "C.co", NA,
  "M",  "R.5",   "R.i",  "Cr.c", "Cl.i", "Cr.5", NA,
  "M",  "R.75",  "R.25", "R.i",  "R.5",  "Cr.c", NA,
  "M",  "R.c",   "R.5",  "R.25", "R.c",  "R.25", "C.i",
  "M",  "A.5",   "R.75", "R.5",  "A.5",  "R.5",  "C.co",
  "M",  "Ac",    "R.c",  "R.75", "Ac",   "R.75", "Cr.5",
  "M",  "Ac",    "A.5",  "R.c",  "Ac",   "R.c",  "Cr.c",
  "M",  "Ac",    "Ac",   "A.5",  "Ac",   "A.5",  "R.i",
  "M",  "Ac",    "Ac",   "Ac",   "Ac",   "Ac",   "R.25",
  "M",  "Ac",    "Ac",   "Ac",   "Ac",   "Ac",   "A.5",
)
```

```{r}
#| label: male-estimates
male_results <- purrr::map_dfr(seq_len(nrow(male)), \(i) {
  est <- estimate_dental_age(male[i, ], verbose = FALSE)
  info <- ac_info(est)
  tibble(
    row = i,
    dental_age = est[["dental_age"]],
    ci_lower = est[["ci_lower"]],
    ci_upper = est[["ci_upper"]],
    n_estimable = info$n_estimable,
    compatibility = info$compatibility
  )
})

male |>
  mutate(row = row_number()) |>
  left_join(male_results, by = "row") |>
  select(-Sex, -row) |>
  mutate(across(c(dental_age, ci_lower, ci_upper), \(x) round(x, 2)))
```

## Notes

- Rows 1--2 use only the canine. With a single tooth, the estimate
  tracks that tooth's reference age directly.
- M1 is added at row 4. When M1 is scored at `C.i`, it is dropped
  because that stage is left-censored in the reference model.
- M3 first appears at row 9. For females, late M3 stages (`R.5`,
  `R.75`, `R.c`, `A.5`) have no published log-normal parameters and
  are excluded.
- From row 11 onward, some teeth reach `Ac` and shift from the age
  estimate to the completion threshold. The `compatibility` column
  shows whether the two agree.
- Row 15 has all teeth `Ac` except M3 at `A.5`. For females, M3 `A.5`
  is unparameterized, so there is no point estimate -- only the
  completion threshold. For males, M3 `A.5` has parameters and
  produces an estimate.
