---
title: "Terminal apex closed stages"
format: html
vignette: >
  %\VignetteIndexEntry{Terminal apex closed stages}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(AgeFromDentition)
```

## Apex closed (A~c~) stages cannot be inverted

The method presented in Šešelj et al. (2019) estimates age by inverting a tooth's stage. Given that a tooth is observed e.g., at `R.5`, Tables 8–13 supply a log-normal distribution for the individual's age.

Apex closure (tooth stage coded `A.c`) breaks that logic, because it is terminal (no exit point). Once a tooth reaches it, it stays there. Observing A~c~ therefore tells you only that the individual is older than the completion age (which has a distribution). It does not localize age further. Tables 8–13 have no A~c~ row for this reason.

For a non-terminal stage $s$ the likelihood is localized:

$$P(s \mid a) = F_{t,s}(a) - F_{t,s+1}(a)$$

The difference $F_{t,s}(a) - F_{t,s+1}(a)$ is the probability of being in stage $s$ at age $a$, having entered $s$ but not yet entered $s+1$. This is the likelihood used to invert the stage into an age estimate.

whereas for the terminal stage it is one-sided, rising with age towards 1:

$$P(\mathrm{A_c} \mid a) = F_{t,\mathrm{A_c}}(a)$$

An A~c~ observation is weakly informative about age, because that observation has an age at attainment (Tables 2-7), which itself is a distribution.

## What we reports instead

Teeth at A~c~ are excluded from the weighted age estimate and used to compute a *reference completion threshold*: a percentile of the sex-specific distribution of age upon *entering* A~c~, taken from Tables 2–7.

```{r}
ac_completion_threshold("F", "M2")$threshold
```

The threshold value represents that 2.5% of the female reference sample had M2 at A~c~ by about 12.2 years.

### What it is not

It is not a lower confidence limit for the individual being assessed. It describes the reference sample, not the person. It is also not a minimum age. Individuals below the threshold exist in the reference sample: 2.5% by definition.

## The all-A~c~ case

When every scored tooth has completed, there is no point estimate:

```{r}
x <- data.frame(
    Sex = "F",
    Canine = "A.c",
    P3 = "A.c",
    P4 = "A.c",
    M1 = "A.c",
    M2 = "A.c",
    M3 = NA
)

estimate_dental_age(x, verbose = FALSE)
```

The threshold still exists, drawn from the five scored teeth:

```{r}
ac_info(estimate_dental_age(x, verbose = FALSE))$threshold
```

An unscored M3 is not treated as completed. Scoring it `A.c`
adds a sixth tooth to the threshold calculation, which changes
the binding tooth:

```{r}
y <- x
y$M3 <- "A.c"

ac_info(estimate_dental_age(y, verbose = FALSE))$threshold
```

## Predictive interval vs. known mean

By default the threshold is computed predictively, widening the
reference standard deviation by the standard error of the fitted mean:

$$\sigma_{\text{eff}} = \sqrt{\mathrm{log\_sd}^2 + \mathrm{se\_log\_mu}^2}$$

Šešelj et al. treat the fitted mean as known. The difference is small for most teeth and important for one:

```{r}
teeth <- c("Canine", "P3", "P4", "M1", "M2", "M3")

data.frame(
    Tooth = teeth,
    predictive = vapply(
        teeth,
        \(t) {
            ac_completion_threshold("F", t)$threshold
        },
        numeric(1)
    ),
    plugin = vapply(
        teeth,
        \(t) {
            ac_completion_threshold("F", t, method = "plugin")$threshold
        },
        numeric(1)
    )
) |>
    transform(difference = round(predictive - plugin, 3))
```

Female M3 moves by almost half a year, because its A~c~ transition is estimated from a single individual. Since M3 attains A~c~ latest of any tooth, it binds the threshold whenever it is scored.

Use `method = "plugin"` if you need the original form. Both `q` and `method` appear in the printed output so that the number can be reproduced from the report alone.

### Uncertainty in standard deviation is not propagated

Uncertainty in the scale parameter (`se_ln_sd` in the source tables, around 0.003–0.004) is *not* propagated. The predictive form is more predictive than plug-in, but not fully predictive.

## Female M3 stage tie

```{r}
StageTies
```

For females, the fitted age of attaining M3 A~c~ is identical to the age of
attaining A~1/2~. Observing apex closure in a female M3 therefore supplies
no information beyond the preceding stage, under this reference model. The
package flags it:

```{r}
info <- ac_info(estimate_dental_age(y, verbose = FALSE))
c(low_precision = info$low_precision, tied = info$tied_transition)
```

Neither flag causes the tooth to be dropped. Discarding the binding tooth
would understate the threshold and throw away the very observation being
reported on.

## Compatibility codes

When some teeth are still developing and others have completed, you get an
estimate, a threshold, and a code relating them:

| Code | Condition |
| :-- | :-- |
| `no_terminal_information` | no tooth at `A.c` |
| `completion_threshold_only` | teeth at `A.c`, none estimable |
| `compatible` | `ci_lower >= threshold` |
| `overlap` | threshold falls inside the interval |
| `discordant` | `ci_upper < threshold` |

The interval is an equal-tailed **central 95%** interval, computed
analytically. It is not a highest-density interval, and it is deliberately
not taken from `estimate_age_hdi()`, which is Monte Carlo and unseeded — a
code built on that would vary between runs on identical input.

`discordant` raises a warning, but does not assert a cause. The
age-given-stage estimator and the attainment threshold come from related
but different models, combined here deliberately approximately, so
disagreement can reflect the method as much as the data.

## Combining several A.c teeth

With more than one completed tooth, the **largest** per-tooth threshold is
reported. This is a transparent reporting convention, not a statistical
combination: it ignores the joint probability of the observed completion
pattern and any correlation between teeth within a person.

The full per-tooth table is always returned, so the convention can be
inspected rather than trusted:

```{r}
ac_completion_threshold("F", c("Canine", "P3", "P4", "M1"))$per_tooth
```

Note that the binding tooth here is P4, not the canine.

## Missing teeth

`NA` means the tooth was not scored, and nothing more. The input format
records presence or absence only, so the package never reports *why* a
tooth is absent — agenesis, extraction, an unscorable image, and a
radiograph that did not include the tooth are analytically different, but
indistinguishable here.

If that distinction matters for your application, record it alongside the
scores; a future release may carry it through.

## Reference

Šešelj M, Sherwood RJ, Konigsberg LW. 2019. Timing of Development of the
Permanent Mandibular Dentition: New Reference Values from the Fels
Longitudinal Study. *Anat Rec* 302:1733–1753.
