---
title: "Terminal (Ac) stages"
format: html
vignette: >
  %\VignetteIndexEntry{Terminal (Ac) stages}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
library(AgeFromDentition)
```

## Why Ac cannot be inverted

The method of Šešelj et al. (2019) estimates age by inverting a tooth's
stage: given that a tooth is observed at, say, `R.5`, Tables 8–13 supply a
log-normal distribution for the individual's age.

Apex closure (`Ac`) breaks that logic, because it is **absorbing**. Once a
tooth reaches it, it stays there for the rest of life. Observing `Ac`
therefore tells you only that the individual is older than the completion
age; it does not localize age at all. Tables 8–13 have no `Ac` row for
precisely this reason.

Formally, for a non-terminal stage $s$ the likelihood is localized:

$$P(s \mid a) = F_{t,s}(a) - F_{t,s+1}(a)$$

whereas for the terminal stage it is one-sided, rising with age towards 1:

$$P(\mathrm{Ac} \mid a) = F_{t,\mathrm{Ac}}(a)$$

An `Ac` observation is weakly informative about age, but it is not
uninformative — so discarding it entirely also throws something away.

## What this package reports instead

Teeth at `Ac` are excluded from the weighted age estimate and used to
compute a **reference completion threshold**: a percentile of the
sex-specific distribution of age at *entering* `Ac`, taken from the
attainment parameters in Tables 2–7.

```{r}
ac_completion_threshold("F", "M2")$threshold
```

That number says: 2.5% of the female reference sample had completed M2 by
about 12.2 years.

### What it is not

It is **not** a 97.5% lower confidence limit for the individual being
assessed. It describes the reference sample, not the person. The package's
output says so explicitly, every time, because the distinction is easy to
lose in a methods section.

It is also not a minimum age. Individuals below the threshold exist in the
reference sample — 2.5% of them, by construction.

## The all-Ac case

When every scored tooth has completed, there is no point estimate at all:

```{r}
x <- data.frame(Sex = "F", Canine = "Ac", P3 = "Ac", P4 = "Ac",
                M1 = "Ac", M2 = "Ac", M3 = NA)

estimate_dental_age(x, verbose = FALSE)
```

A tempting shortcut is to return a terminal age — 18, say — in this
situation. The package does not, and will not. That value is unsupported by
the reference data, and it maps very different completion patterns onto one
number: an individual with only M1 complete and one with M3 complete would
receive the same age.

Note also that an unscored M3 is never treated as a completed one:

```{r}
y <- x
y$M3 <- "Ac"

ac_info(estimate_dental_age(y, verbose = FALSE))$threshold
ac_info(estimate_dental_age(x, verbose = FALSE))$threshold
```

## Predictive versus plug-in

By default the threshold is computed **predictively**, widening the
reference SD by the standard error of the fitted mean:

$$\sigma_{\text{eff}} = \sqrt{\mathrm{log\_sd}^2 + \mathrm{se\_log\_mu}^2}$$

The source paper's form treats the fitted mean as known. The difference is
immaterial for most teeth and decisive for one:

```{r}
teeth <- c("Canine", "P3", "P4", "M1", "M2", "M3")

data.frame(
  Tooth = teeth,
  predictive = vapply(teeth, \(t) {
    ac_completion_threshold("F", t)$threshold
  }, numeric(1)),
  plugin = vapply(teeth, \(t) {
    ac_completion_threshold("F", t, method = "plugin")$threshold
  }, numeric(1))
) |>
  transform(difference = round(predictive - plugin, 3))
```

Female M3 moves by almost half a year, because its `Ac` transition is
estimated from a **single individual**. Since M3 attains `Ac` latest of any
tooth, it binds the threshold whenever it is scored — so the two
conventions disagree most in exactly the case this feature exists to serve.

Use `method = "plugin"` if you need the source paper's form. Whichever you
choose, declare it: both `q` and `method` appear in the printed output so
that the number can be reproduced from the report alone.

### A known incompleteness

Uncertainty in the scale parameter (`se_ln_sd` in the source tables, around
0.003–0.004) is *not* propagated. The predictive form is more predictive
than plug-in, not fully predictive.

## The female M3 tie

```{r}
StageTies
```

For girls, the fitted age of attaining M3 `Ac` is identical to the age of
attaining `A.5`. Observing apex closure in a female M3 therefore supplies
no information beyond the preceding stage, under this reference model. The
package flags it:

```{r}
info <- ac_info(estimate_dental_age(y, verbose = FALSE))
c(low_precision = info$low_precision, tied = info$tied_transition)
```

Neither flag causes the tooth to be dropped. Discarding the binding tooth
would understate the threshold and throw away the very observation being
reported on.

## Compatibility codes

When some teeth are still developing and others have completed, you get an
estimate, a threshold, and a code relating them:

| Code | Condition |
| :-- | :-- |
| `no_terminal_information` | no tooth at `Ac` |
| `completion_threshold_only` | teeth at `Ac`, none estimable |
| `compatible` | `ci_lower >= threshold` |
| `overlap` | threshold falls inside the interval |
| `discordant` | `ci_upper < threshold` |

The interval is an equal-tailed **central 95%** interval, computed
analytically. It is not a highest-density interval, and it is deliberately
not taken from `estimate_age_hdi()`, which is Monte Carlo and unseeded — a
code built on that would vary between runs on identical input.

`discordant` raises a warning, but does not assert a cause. The
age-given-stage estimator and the attainment threshold come from related
but different models, combined here deliberately approximately, so
disagreement can reflect the method as much as the data.

## Combining several Ac teeth

With more than one completed tooth, the **largest** per-tooth threshold is
reported. This is a transparent reporting convention, not a statistical
combination: it ignores the joint probability of the observed completion
pattern and any correlation between teeth within a person.

The full per-tooth table is always returned, so the convention can be
inspected rather than trusted:

```{r}
ac_completion_threshold("F", c("Canine", "P3", "P4", "M1"))$per_tooth
```

Note that the binding tooth here is P4, not the canine.

## For your methods section

Adapt the following. It states the four things that make the result
reproducible: the reference values, the treatment of `Ac`, the convention
that produced the threshold, and what happens when nothing is estimable.

> Dental age was estimated with `AgeFromDentition` version 0.2.0, using
> the sex-specific reference values of Šešelj et al. (2019).
> Non-terminal stages were combined by precision-weighted averaging of the
> age-given-stage log-normal approximations (Tables 8–13). The terminal
> stage Ac was treated as right-censored and excluded from that estimate.
> For each tooth scored Ac, a reference completion threshold was
> calculated from the age-of-attainment parameters (Tables 2–7) as
> `exp(theta_Ac + z_q * sqrt(lnSD^2 + se_theta^2))` with `q = 0.025`, the
> 2.5th percentile of the predictive distribution of age at attaining Ac
> for an individual from the reference sample; where several teeth were
> scored Ac, the largest such threshold is reported. This threshold
> describes the reference sample and is not a lower confidence limit for
> the individual. Where no non-terminal tooth was scorable, no point
> estimate was produced and only the threshold is reported.

If you used `method = "plugin"`, replace the formula sentence with:

> ... calculated from the age-of-attainment parameters (Tables 2–7) as
> `exp(theta_Ac + z_q * lnSD)` with `q = 0.025`, the 2.5th percentile of
> the fitted reference distribution of age at attaining Ac.

## Missing teeth

`NA` means the tooth was not scored, and nothing more. The input format
records presence or absence only, so the package never reports *why* a
tooth is absent — agenesis, extraction, an unscorable image, and a
radiograph that did not include the tooth are analytically different, but
indistinguishable here.

If that distinction matters for your application, record it alongside the
scores; a future release may carry it through.

## Reference

Šešelj M, Sherwood RJ, Konigsberg LW. 2019. Timing of Development of the
Permanent Mandibular Dentition: New Reference Values from the Fels
Longitudinal Study. *Anat Rec* 302:1733–1753.
